---
title: "Basic ENMTML usage"
authors: "André Andrade, Santiago Velazco & Paulo De Marco"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic ENMTML usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=F
)
```

## Basic *ENMTML* Run

We present here a simple model, with the simplest parameters and decisions.  
The main objective of this run is to introduce the main arguments and help with some details which are required to run the function.   
**We will use the files available at the package example, so that everyone can easily reproduce those steps!**

## Creating the folder structure from the example files

**ENMTML** example data set consists of five bioclimantic variables for the current period and five different virtual species.  

In order to simulate the files and folders needed for an standard **ENMTML** run we will create different folders containing the example data.  

For this we will use the *raster* package:

```{r include=F}
load("C:/Users/decoa/OneDrive/Trabajos/22 - ENM_Package/ENMTML/Data/occ.RData")
load("C:/Users/decoa/OneDrive/Trabajos/22 - ENM_Package/ENMTML/Data/env.RData")
```

```{r}
if (!"raster"%in%installed.packages()){install.packages("raster")}  
require(raster)  


# First we will create a folder (ENMTML_example) within the working directory
getwd() #' Working directory of R session
d_ex <- file.path(getwd(), 'ENMTML_example')
d_ex
dir.create(d_ex)

# We will now save ENMTML data set within the ENMTML_example folder

# Virtual species occurrences
data("occ")

# See the format of the .TXT file(tab-separated)! 
# We only need three columns:
  # one containing species name
  # one containing longitude information 
  # one containing latitude information

knitr::kable(occ[c(1:4,43:46,90:94,127:130),])
d_occ <- file.path(d_ex, 'occ.txt')
utils::write.table(occ, d_occ, sep = '\t', row.names = FALSE)

# Five bioclimatic variables for current conditions
data("env")
d_env <- file.path(d_ex, 'current_env_var')
dir.create(d_env)
raster::writeRaster(env, file.path(d_env, names(env)), bylayer=TRUE, 
                    overwrite=T, format='GTiff')

# shell.exec(d_ex) # open the directory and folders created
rm(list = c('env','occ'))

# We now we have the minimum data required to create models with ENMTML package!
# A directory with environmental rasters and a .txt file with occurrence.

```


**The following objects contains the path to the folder with predictors and the path to the occurrence file (TXT)**

* **d_occ** : File path to species occurrence file (.txt; tab-separated)
* **d_env** : Directory path with current environmental conditions (here in .tiff)


## Model design  

We will now fit models for with the following specifications: 

* **Five** different algorithms: Bioclim, Generalized Linear Model, Random Forests, Support Vector Machine and Maxent  
* **Random** pseudo-absence allocation  
* **No** accessible area restriction  
* **Random** partition between training-test datasets  
* **No** Ensemble  
* **No** Bias correction  
* **No** Collinearity control in the predictors  
* **Binary maps** will be produced using the *Maximum Specificity and Sensitivity Threshold*  


```
require(ENMTML)

ENMTML(
 pred_dir = d_env,
 proj_dir = NULL,
 result_dir = file.path(d_ex,"Result"),
 occ_file = d_occ,
 sp = 'species',
 x = 'x',
 y = 'y',
 min_occ = 10,
 thin_occ = NULL,
 eval_occ = NULL,
 colin_var = NULL,
 imp_var = FALSE,
 sp_accessible_area = NULL,
 pseudoabs_method = c(method = 'RND'),
 pres_abs_ratio = 1,
 part=c(method= 'KFOLD', folds='2'),
 save_part = FALSE,
 save_final = TRUE,
 algorithm = c('BIO','GLM', 'RDF', 'SVM', 'MXD'),
 thr = c(type='MAX_TSS'),
 msdm = NULL,
 ensemble = NULL,
 extrapolation = FALSE,
 cores = 1
)
```


### Results

**ENMTML** function will create a folder named **Result** within the *ENMTML_example* folder containing the *Algorithms* sub-folder and several  *.txt* files:  

* **Evaluation_Table.txt** Contains the results for model evaluation, with several metrics    
* **InfoModeling.txt** Information of the chosen modeling parameters       
* **Number_Unique_Occurrences.txt** Number of unique occurrences for each species    
* **Occurrences_Cleaned.txt** Dataset produced after selecting a single occurrence per grid-cell(*unique occurrences*)       
* **Thresholds_Algorithm.txt** Information about the thresholds used to create the presence-absence maps for each algorithm (Presence-absence maps are created from the Threshold of complete models)    
* **Moran & Mess** Contains information about autocorrelation and environmental similatiry between the datasets used to fit and evaluate the model  


## What's next?

We will see next several methodological advances for the Standard model and how those are incorpoated in our modelling routine.  
We will also cover projection for other periods, extent and include other algorithms and produce and ensemble model!

__Hope you understood the logic behind **ENMTML** and were able to produce your first models!__  

**See you next time!**

> Feel free to contact us by mail (you can find the paths to André and Santiago e-mails at the end of the [GitHub page](https://github.com/andrefaa/ENMTML))


